### 2011-07-20 Paper/Paper.tex
Updated paper. Final tweaks to the system

% Fancy maths stuff
\usepackage{amsmath, amsthm, amssymb}

@@

@@

NOTE(Show pitch and onset trees here)

To translate a delta tree into a segmentation we simply take all the nodes at a particular level and consider the notes it yields as one group. However, as can be seen in figure \ref{pitchonsettrees} the trees are quite deep and it is not clear what level we should choose. We can make the trees shallower by slightly modifying the delta rule. We introduce a small tolerance parameter $\epsilon$ and instead of looking for the maximum in of a list of deltas, we look for deltas within $m - \epsilon$ and $m + \epsilon$, where $m$ is the maximum delta within a list of deltas.

Using this modification of the delta rule we can bring the delta onset tree back a shallow tree in which it is safe to choose groupings defined by the first level nodes. These groupings are safe, but also quite large. Some music, like Bach's inventionen uses almost no differentiation in IOI. Hence, a segmentation based solely on inter onset interval will not serve our purposes.

To overcome this problem it we will use pitch. While Bach's inventionen use almost no differentiation in IOI they can be segmented nonetheless. We will not make any attempt to combine pitch interval trees and IOI trees. Instead we take the segmentation given by the IOI tree and try to subdivide every segment using a pitch interval tree.

@@

@@

\subsection{Dealing with polyphony}
\begin{itemize}
\item Use only melody
\end{itemize}
\subsection{Discretization}
To calculate probabilities, we must discretize the feature and parameter vectors.

@@

To discretize the features, we simply normalize every feature dividing through the maximum absolute value of that feature found in the corpus. After normalization we multiply the feature by a discretization parameter that determines the number of bins and take the floor.

@@

\subsection{Performance rendering}
\begin{itemize}
\item Extract melody
\item Extract scorefeatures from melody
\item lookup notes in deviation, ignore missing notes
\item Extract expressive and non-expressive melody and extract expression features
\item Perform notes other than melody notes the same as the nearest melody note
\end{itemize}

\section{Evaluation and Results}
The proposed system actually requires more training data to do meaningful statistics. However, since expressive interpretation can be 
\subsection{Subjective Listening}
\subsection{Correlation}
\section{Integration with Note Level Performance Rendering}

@@

\section{Conclusion}
We have introduced structure level expression and argued that a good performance rendering systems needs some notion of structure level expression.

@@

\begin{itemize}
\item Top notes is not always the melody, musical attention
\item Bass and harmony shouldn't be played with the same expression as melodynotes
\item Dataset size
\item Targets for expression level expression
\item Expressive markings
\item Staccato/articulation statistics
\end{itemize}

@@

>>>>

% Fancy maths stuff
\usepackage{amsmath, amsthm, amssymb}
%\usepackage{fullpage}

@@

% Images
\usepackage{graphicx}

@@

\begin{figure}

\end{figure}
NOTE(we now use only onset on the mozart corpus, explain adaptive segmentation)

In the tree in figure \ref{pitchonsettrees} we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes. To translate the tree into a segmentation we do the following for for every node in the tree starting with the root node:
\begin{algorithm}
\label{alg:segmentation}
\begin{enumerate}
\item If the node is a note, at it as a singleton constituent to the segmentation
\item Expand the node.
\item If the node contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
\item If the node contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
\end{enumerate}
\end{algorithm}

The value of $x$ represents the tolerance of singleton constituents. We need this parameter because sometimes very long notes cause top level splits in the onset tree. We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart.

%To translate a delta tree into a segmentation we simply take all the nodes at a particular level and consider the notes it yields as one group. However, as can be seen in figure \ref{pitchonsettrees} the trees are quite deep and it is not clear what level we should choose. We can make the trees shallower by slightly modifying the delta rule. We introduce a small tolerance parameter $\epsilon$ and instead of looking for the maximum in of a list of deltas, we look for deltas within $m - \epsilon$ and $m + \epsilon$, where $m$ is the maximum delta within a list of deltas.

%Using this modification of the delta rule we can bring the delta onset tree back a shallow tree in which it is safe to choose groupings defined by the first level nodes.

For some music like Mozart, using only a delta onset tree works very well. However, some music, like Bach's inventionen uses almost no differentiation in IOI. Hence, a segmentation based solely on inter onset interval will not serve our purposes for this music.

To overcome this problem we will use pitch. While Bach's inventionen use almost no differentiation in IOI they can be segmented nonetheless. We will not make any attempt to combine pitch interval trees and IOI trees. Instead we take the segmentation given by the IOI tree and try to subdivide every segment using a pitch interval tree.

@@

\begin{figure}
\includegraphics[scale=3]{img/melodysegments}
\caption{First four segments of piano sonata KV331 III by Mozart (grace notes are not shown)}
\end{figure}

@@

\subsection{Discretization}
To calculate probabilities, we must discretize the feature and parameter vectors.

@@

To discretize the features, we simply normalize every feature dividing through the maximum absolute value of that feature found in the corpus. After normalization we multiply the feature by a discretization parameter $d$ that determines the number of bins and take the floor.

@@

\subsection{Smoothing}

Despite having discretized our feature and observation vectors we often find that we observe feature vectors in a new score that we had never seen during training. We do not want the conditional probabilities from equation \ref{emission} to become zero so we have to smooth these probabilities. We use a smoothing technique known as simple Good-Turing smoothing. The idea is that we use the probabilities of things we have seen once for the things we have never seen. Recall how we calculate the conditional probability of a feature vector:
\[
P(f_i|e_i) = \frac{c(f_i, e_i)}{c(e_i)}
\]

Let us call the $c(f_i, e_i)$ the coincedence count. Let $N_c$ be the number of things with frequency $c$ in the corpus. For example if there are five coincedences $(f_x, e_x)$ nd no other pair of $f$ and $e$ occurs five times, then $N_{5} = 20$. Good-Turing re\"estimates counts according to this formula:

\begin{equation}
\label{count}
c^* = (c+1) \frac{N_{c+1}}{N_c}
\end{equation}

The smoothed probability of some event $x$ is

\[P(x) = \frac{c^*}{N}\]

In our case, the probability we seek is $P(f_i|e_i)$, the sample size $N$ is therefore the number of times we have seen $e_i$: $c(e_i)$. $N_c$ is the number of coincedences with $e_i$ that we have seen $c$ times. The count $c$ is $c(f_i, e_i)$ and $c^*$ is derived using equation \ref{count}. The

\[P(f_i|e_i) = \frac{c^*}{c(e_i)} \mbox{ for all } c > 0 \]

By applying this formula to all counts larger than zero we reserve approximately $\frac{N_1}{N}$ probability mass for things that we have never seen. We assign an equal probability for all unseen feature vectors, namely:
\[
P(f_{\texttt{unseen}}|e_i) = \frac{1}{U} \frac{N_1}{c(e_i)} 
\]
Where U is the number of unseen coincedences, which is determined by the number of different observations in the corpus plus the new observations from the score minus the number of coincedences with $e_i$.

In order to make this work we cannot use $N_c$ directly since it will not be defined for every $c$. We use a least square apprixmation using the following function.

\[\log(N_c) = a + b \log(c)\]

This completes the definition of simple Good-Turing smoothing. However, even now there may still be some expression parameter vectors that occur with unique feature vectors, so for that expression only $N_1$ is defined. We cannot fit a function to one sample, so if we have only one $Nc$ sample, we simply turn off Good-Turing smoothing and accept the fact that some probabilities will be zero.

\subsection{Performance rendering}

Our model is able to generate expressively performed melodies but does not handle polyphony. During training, the bass and harmony notes were stripped off. After rendering an expressive performance we can simply put them back in and estimate their expressive parameters. We do this by this by by giving each  bass and harmony note the expressive parameters of the last played melody note.

\begin{itemize}
\item Extract melody
\item Extract scorefeatures from melody
\item lookup notes in deviation, ignore missing notes
\item Extract expressive and non-expressive melody and extract expression features
\item Perform notes other than melody notes the same as the nearest melody note
\end{itemize}

\section{Evaluation and Results}

Using the method that we now have extensively described we have tried to generate a few performances of piano works by Mozart. The system was trained on 42 performances of 13 works by Mozart, see appendix \ref{appendix:corpus} for a more details. The choice for Mozart was motivated by the fact that our segmentation works best when phrase transitions are marked by long notes or silences, which is often the case in the work of Mozart. Another reason for choosing Mozart in favor of for example Chopin is that we think structure level expression is more outspoken in performances of Mozart's music.

%The system was tested on works from Mozart as we do not expect our system to generalize well to other composers.  We used

We discretized the score features using a discretization parameter of five. This means the constituent features (see section \ref{sec:features}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well. The expression features where discretized into ten bins, with a sensitivity of five. These settings made sure we recognised most of the score features of a new score. The low number of bins does give the performances bit of a cartoonistic quality, but that does not make them any less interesting.

\subsection{Subjective Listening}
\subsection{Correlation}
\section{Integration with Note Level Performance Rendering}

@@

\section{Conclusion}
In this thesis we have critisized performance rendering systems that only use note level expression.

We have introduced structure level expression and argued that a good performance rendering systems needs some notion of structure level expression.

@@

\begin{itemize}
\item Top notes is not always the melody, musical attention
\item Bass and harmony shouldn't be played with the same expression as melodynotes
\item Dataset size
\item Targets for expression level expression
\item Expressive markings
\item Staccato/articulation statistics
\item Backoff smoothing (doesn't assign equal probabilities to unseen events)
\end{itemize}

@@

\appendix
\section{Corpus}
\label{appendix:corpus}
\begin{tabular}{ll}
\textbf{Work} & \textbf{Performer}\\
\hline
Sonata KV331 \MakeUppercase{\romannumeral 1}. & Hiroko Nakamura\\
& Glenn Gould\\
& Christoph Eschenbach\\
& Ingrid Haebler\\
& Lili Kraus\\
& Maria Jo\~ao Pires\\
& Alicia De Larrocha\\
& kn?\\
& yi?\\
& tn?\\
& Norio Shimizu\\
& mo\\
& mk?\\
& ea?\\
& nm?\\
& kt?\\
& tm?\\
\hline
Sonata KV331 \MakeUppercase{\romannumeral 2}. & Hiroko Nakamura\\
& Maria Jo\~ao Pires\\
& Alicia De Larrocha\\
& ea?\\
\hline
Sonata KV331 \MakeUppercase{\romannumeral 3}. & Hiroko Nakamura\\
& Maria Jo\~ao Pires\\
& Christoph Eschenbach\\
\hline
Sonata KV545 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
& Glenn Gould\\
\hline
Sonata KV545 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
& Glenn Gould\\
\hline
Sonata KV545 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
& Glenn Gould\\
\hline
Sonata KV279 \MakeUppercase{\romannumeral 1}. & Glenn Gould\\
& Maria Jo\~ao Pires\\
\hline
Sonata KV279 \MakeUppercase{\romannumeral 2}. & Glenn Gould\\
& Maria Jo\~ao Pires\\
\hline
Sonata KV279 \MakeUppercase{\romannumeral 3}. & Glenn Gould\\
& Maria Jo\~ao Pires\\
\hline
Sonata KV310 \MakeUppercase{\romannumeral 1}. & Maria Jo\~ao Pires\\
& Glenn Gould\\
& Hiroko Nakamura\\
Sonata KV310 \MakeUppercase{\romannumeral 2}. & Maria Jo\~ao Pires\\
Sonata KV310 \MakeUppercase{\romannumeral 3}. & Maria Jo\~ao Pires\\
Sonata KV570 \MakeUppercase{\romannumeral 3}. & nm?\\
\end{tabular}
